---
title: "FR07"
author: "Li-Fang Yeo"
date: "2026-05-12"
output: html_document
---

---
  title: "Obesity indicator - EASTWEST"
author: "Li-Fang Yeo"
date: "06-June-2025"
output: html_document
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```
```{r}
library(magrittr)
library(ANCOMBC)
library(mia)
library(ggplot2)
library(dplyr)

devtools::load_all()
```
Command line arguments
```{r}
args <- list(
  optparse::make_option("--east", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--west", action="store_true", default=FALSE, help="Exclude Western Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--low", action="store_true", default=FALSE, help="Exclude low GFR [default \"%default\"]"),
  optparse::make_option("--high", action="store_true", default=FALSE, help="Exclude high GFR [default \"%default\"]"),
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
  optparse::make_option("--prevalence", type = "numeric", default = 5/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()

args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```
Data definition
```{r}
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             VYOTARO = "Waist circ",
             WHR = "waist-hip ratio",
             PREVAL_DIAB = "Diabetes",
             PREVAL_CVD = "Prevalent cardiovascular disease",
             CURR_SMOKE = "Smoking",
             ALKI2_FR07 = "Alcohol consumption per week (g)",
             #HFC = "Healthy food choices",
             Q57X = "Exercise",
             shannon = "Shannon diversity",
             observed = "Observed richness",
             EAST = "Eastern finland")
             #total_reads = "Total reads"
```
Read data in and filter for covariates
```{r}
tse07 <- readRDS("../data/tse_mgs-20241118_104759.rds") %>%
  tse_filter(PROJECT =="FR07") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::addAlpha(assay.type = "counts", index = c("shannon", "observed"), name = c("shannon", "observed")) %>%
  #tse_add_food_score(HFC) %>%
  tse_mutate(dplyr::across(c(MEN,
                             CURR_SMOKE,
                             EAST,
                             Q57X,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  tse_mutate(dplyr::across(c(shannon,
                             observed), as.numeric)) %>%
  #pregnant
  tse_filter(GRAVID %in% c(1, NA)) %>%
  #antibiotic use in the past 1 month
  tse_filter(BL_USE_RX_J01_1mo %in% c(0, NA)) %>%
  #missing covariate
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars)), not_na)) %>%
  #low reads
  #tse_filter(total_reads > 50000) %>%
  #filter for covariates
  tse_select(names(vars)) 
```

# Characteristic
```{r}
tse07 %>%
  tse_meta(rownames = FALSE) %>%
  #dplyr::mutate(MEN = factor(ifelse(MEN == 1, "MEN" , "WOMEN"))) %>%
  #dplyr::mutate(EAST = factor(ifelse(EAST == 1, "Discovery east" , "Validation west"))) %>%
  mytableone(vars) %>%
  write.csv("tableone_fr07.csv")
```
```{r}
#manual calculation for total_reads and observed richness
as.data.frame(colData(tse)) %>%
  #filter(EAST == 0) %>%
  summarise(
    mean_total_reads = mean(total_reads, na.rm = TRUE),
    sd_total_reads   = sd(total_reads, na.rm = TRUE),
    mean_observed    = mean(observed, na.rm = TRUE),
    sd_observed      = sd(observed, na.rm = TRUE)
  )


```

# Model function definitions
BMI model + covariates
```{r}
lm_bmi_partial <-  purrr::partial(lm,
                                  formula = scale(BMI) ~
                                    scale(term) + BL_AGE + MEN + CURR_SMOKE +
                                    ALKI2_FR07 +  Q57X)

```
```{r}
lm_bmi_model_for_taxon <- function(df, term, rank = NULL) {
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    lm_bmi_partial(data = .)
}
```
# Alpha diversity using microbial counts on Shannon index
## Univariable adjusted for age and sex
BMI
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  lm(scale(BMI) ~ scale(shannon) + scale(BL_AGE) + MEN + CURR_SMOKE +
       scale(ALKI2_FR07)  + Q57X, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity to BMI")
```
```{r}
#observed richness
tse %>%
  tse_meta(rownames = FALSE) %>% 
  lm(scale(BMI) ~ scale(observed) + scale(BL_AGE) + MEN +  CURR_SMOKE +
       scale(ALKI2_FR07)  + Q57X, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity to BMI")
```

Significant taxa from internal validation cohort
```{r}
#df_lm comes from the discovery cohort
sig.list <- read.delim("C:/Users/lifyeo/bmi/df_lm_bmi.tsv") %>%
  dplyr::mutate(across(everything(),~ gsub("GUT_","",.))) %>%
  as.data.frame()
list <- sig.list$taxa
list
#subset tse by sig list
selected_rows <- rowData(tse07)$Species %in% list &
  !is.na(rowData(tse07)$Species)
tse_val <- tse07[selected_rows, ]
```
#Common taxa
Agglomerate prevalent features
```{r}
#reads were filtered to detection threshold of 0.1% relative abundance, in 5% of samples
#This is to filter out rare taxa, to ensure we are anlaysing microbes that are likely to be real and not sequencing artefacts
subset_features_partial <-  purrr::partial(mia::agglomerateByPrevalence,
                                           x = tse_val,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
Subset to only Species
```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
Melt tse to data frame
```{r}
melt_tse <- function(x, method = "clr") {
  mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltSE(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
## Waist to hip ratio
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_bmi_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_bmi_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) 
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_bmi_results %>%
  dplyr::arrange(qval_fdr) %>% 
  dplyr::filter(p.value < 0.05) %>%
  write.csv("df_lm_bmi_FR07_p.csv")
  DT::datatable(caption = "Linear model for BMI FR07 validation")
```
#Functional validation
```{r}
joined_pathabundance <- read_delim("../data/joined_pathabundance.tsv", 
                                   delim = "\t", escape_double = FALSE, trim_ws = TRUE)
colnames(joined_pathabundance) <- gsub("\\.R1.*","", colnames(joined_pathabundance))
colnames(joined_pathabundance) <- ifelse(
  startsWith(colnames(joined_pathabundance), "40"),
  gsub("-", "", colnames(joined_pathabundance)),
  colnames(joined_pathabundance)
)
common_ids <- intersect(colnames(joined_pathabundance), rownames(colData(tse07)))


shit <- joined_pathabundance %>%
  select(all_of(c("Pathway", common_ids))) %>% 
  filter(rowSums(across(where(is.numeric))) != 0) %>%
  #removes all these pathways
  filter(!grepl("\\|g_", Pathway)) %>%
  filter(!grepl("superpathway", Pathway)) %>%
  filter(!grepl("fungi", Pathway)) %>%
  filter(!grepl("plant", Pathway)) %>%
  filter(!grepl("eukaryotes", Pathway)) %>%
  filter(!grepl("yeast", Pathway))

shit.t <- as.data.frame(t(shit))
#take first row as colname
colnames(shit.t) <- shit.t[1,]
#remove extra row
shit.t <- shit.t[-1, ] 
#change from character to numeric
shit.t <- shit.t %>% mutate_if(is.character, as.numeric) %>%
  tibble::rownames_to_column("Barcode") 

#You will get a dataframe of prevalent pathways
#poo needs to have samples as row, pathways as column
shit.10 <- shit.t %>% 
  select(-1) %>%
  #mutate to 1 for everything > 0.0000000001
  mutate(across(everything(), ~ . > 1e-10)) %>%
  #get proportion
  summarize(across(everything(), mean)) %>%
  tidyr::gather(Pathway,prevalence) %>%
  #filter for pathways that are prevalent in 5%
  dplyr::filter(prevalence >0.05) 

# filtered to only prevalent pathways
hi <- shit.10 %>% 
  select(-2) %>%
  dplyr::inner_join(shit, by="Pathway" ) %>%
  t() %>%
  as.data.frame()

#take first row as colname
colnames(hi) <- hi[1,]
#remove extra row
hi <- hi[-1, ] 
#change from character to numeric
hi <- hi %>% mutate_if(is.character, as.numeric) %>%
  tibble::rownames_to_column("Barcode") 

transform_column <- function(x) {
  qnorm((rank(x, na.last = "keep") - 0.5) / sum(!is.na(x)))
}

# Apply the transformation to all columns
poo_rank <- hi
poo_rank[, -1] <- lapply(hi[, -1], transform_column)
```

```{r}
lm_full_bmi_partial <-  purrr::partial(lm,
                                  formula = scale(BMI) ~
                                    scale(term) + BL_AGE + MEN +
                                    CURR_SMOKE + ALKI2_FR07 + Q57X) %>%
                    purrr::possibly()

FR07 <- as.data.frame(colData(tse07))%>%
  tibble::rownames_to_column("Barcode")

#Here you will have the correct number of samples
df <- dplyr::right_join(poo_rank, FR07, by = "Barcode")

#take only significant pathway in discovery cohort
sig <- read.csv("C:/Users/lifyeo/bmi/SigPathway.csv") %>%
  select(Pathway)

```
```{r}
##Linear model for inverse rank
#Take the column names from Pathway in shit dataframe as a list and mutate
only_sig <- sig %>%
  dplyr::mutate(results = purrr::map(Pathway, ~lm_full_bmi_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

only_sig %>% 
  dplyr::filter(!is.na(results)) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "scale(term)") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  #dplyr::filter(qval_fdr < 0.05) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  write.csv("functional-IRN-FR07.csv")
  DT::datatable(caption = "Linear model for Inverse Rank MetaCyc pathways - validation west, multi-adj")

