---
title: "Obesity indicator - EASTWEST"
author: "Li-Fang Yeo"
date: "06-June-2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```
```{r}
library(magrittr)
library(ANCOMBC)
library(mia)
library(ggplot2)
library(dplyr)
library(maaslin3)

devtools::load_all()
```
Command line arguments
```{r}
args <- list(
  optparse::make_option("--east", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--west", action="store_true", default=TRUE, help="Exclude Western Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--low", action="store_true", default=FALSE, help="Exclude low GFR [default \"%default\"]"),
  optparse::make_option("--high", action="store_true", default=FALSE, help="Exclude high GFR [default \"%default\"]"),
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
optparse::make_option("--prevalence", type = "numeric", default = 5/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()

args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```
Data definition
```{r}
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             VYOTARO = "Waist circ",
             WHR = "waist-hip ratio",
             PREVAL_DIAB = "Diabetes",
             PREVAL_CVD = "Prevalent cardiovascular disease",
             CURR_SMOKE = "Smoking",
             ALKI2_FR02 = "Alcohol consumption per week (g)",
             HFC = "Healthy food choices",
             Q57X = "Exercise",
             shannon = "Shannon diversity",
             observed = "Observed richness",
             EAST = "Eastern finland",
             total_reads = "Total reads")
```
Read data in and filter for covariates
```{r}
tse <- readRDS("./data/tse_mgs-20241118_104759.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::addAlpha(assay.type = "counts", index = c("shannon", "observed"), name = c("shannon", "observed")) %>%
  tse_add_food_score(HFC) %>%
  tse_mutate(dplyr::across(c(MEN,
                             CURR_SMOKE,
                             EAST,
                             Q57X,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  #pregnant
  tse_filter(GRAVID %in% c(1, NA)) %>%
  #antibiotic use in the past 1 month
  tse_filter(BL_USE_RX_J01_1mo %in% c(0, NA)) %>%
  #missing covariate
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars)), not_na)) %>%
  #low reads
  tse_filter(total_reads > 50000) %>%
  #filter for covariates
  tse_select(names(vars))

#turn on east/west
tse1 <- tse %>%
    #filter for eastern finland
  { if (args$east) tse_filter(., EAST == 0) else . } %>% 
  { if (args$west) tse_filter(., EAST == 1) else . } %>% 
  tse_select(names(vars))

```

# Characteristic
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  #dplyr::mutate(MEN = factor(ifelse(MEN == 1, "MEN" , "WOMEN"))) %>%
  dplyr::mutate(EAST = factor(ifelse(EAST == 1, "Discovery east" , "Validation west"))) %>%
  mytableone(vars,fo =  ~ .| EAST ) 
```
# Model function definitions
BMI model + covariates
```{r}
lm_bmi_partial <-  purrr::partial(lm,
                                  formula = scale(BMI) ~
                                    scale(term) + BL_AGE + MEN +
                                    PREVAL_DIAB + PREVAL_CVD + CURR_SMOKE +
                                    ALKI2_FR02 + HFC + Q57X)
```
```{r}
lm_bmi_model_for_taxon <- function(df, term, rank = NULL) {
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    lm_bmi_partial(data = .)
}
```
WHR model + covariates
```{r}
lm_whr_partial <-  purrr::partial(lm,
                                  formula = scale(WHR) ~
                                    scale(term) + BL_AGE + MEN +
                                    PREVAL_DIAB + PREVAL_CVD + CURR_SMOKE +
                                    ALKI2_FR02 + HFC + Q57X)
```
```{r}
lm_whr_model_for_taxon <- function(df, term, rank = NULL) {
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    lm_whr_partial(data = .)
}
```
# Alpha diversity using microbial counts on Shannon index
## Univariable adjusted for age and sex
BMI
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  lm(scale(BMI) ~ scale(shannon) + scale(BL_AGE) + MEN + PREVAL_DIAB + PREVAL_CVD + CURR_SMOKE +
                                    scale(ALKI2_FR02) + scale(HFC) + Q57X, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity to BMI")
```
```{r}
#observed richness
tse %>%
  tse_meta(rownames = FALSE) %>% 
  lm(scale(BMI) ~ scale(observed) + scale(BL_AGE) + MEN + PREVAL_DIAB + PREVAL_CVD + CURR_SMOKE +
                                    scale(ALKI2_FR02) + scale(HFC) + Q57X, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity to BMI")
```
Hip to waist ratio
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  lm(scale(WHR) ~ scale(shannon) + scale(BL_AGE) + MEN + PREVAL_DIAB + PREVAL_CVD + CURR_SMOKE +
                                    scale(ALKI2_FR02) + scale(HFC) + Q57X, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity to WHR")
```
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  lm(scale(WHR) ~ scale(observed) + scale(BL_AGE) + MEN + PREVAL_DIAB + PREVAL_CVD + CURR_SMOKE +
                                    scale(ALKI2_FR02) + scale(HFC) + Q57X, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  write.csv("alpha-whr-rich.csv")
  DT::datatable(caption = "Alpha diversity to WHR")
```
#Common taxa
Agglomerate prevalent features
```{r}
#reads were filtered to detection threshold of 0.1% relative abundance, in 5% of samples
#This is to filter out rare taxa, to ensure we are anlaysing microbes that are likely to be real and not sequencing artefacts
subset_features_partial <-  purrr::partial(mia::agglomerateByPrevalence,
                                           x = tse1,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
Subset to only Species
```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
Melt tse to data frame
```{r}
melt_tse <- function(x, method = "clr") {
  mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltSE(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)

```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
## BMI
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_bmi_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_bmi_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_bmi_results %>%
  dplyr::arrange(qval_fdr) %>% 
  #write.csv("df_bmi.csv")
  DT::datatable(caption = "Linear model for BMI - discovery east")
```
## WaistCirc
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_waist_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_waist_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_waist_results %>%
  dplyr::arrange(qval_fdr) %>%
  DT::datatable(caption = "Linear model for Waist Circ")
  
```
## Waist to hip ratio
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_whr_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_whr_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_whr_results %>%
  dplyr::arrange(qval_fdr) %>% 
  #dplyr::filter(qval_fdr < 0.05) %>%
  DT::datatable(caption = "Linear model for Waist Hip ratio")
```
Command line arguments
```{r}
args <- list(
  optparse::make_option("--east", action="store_true", default=TRUE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--west", action="store_true", default=FALSE, help="Exclude Western Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--low", action="store_true", default=FALSE, help="Exclude low GFR [default \"%default\"]"),
  optparse::make_option("--high", action="store_true", default=FALSE, help="Exclude high GFR [default \"%default\"]"),
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
optparse::make_option("--prevalence", type = "numeric", default = 5/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()
```
```{r}
args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```
Read data in and filter for covariates
```{r}
#turn on east/west
tse0 <- tse %>%
    #filter for eastern finland
  { if (args$east) tse_filter(., EAST == 0) else . } %>% 
  { if (args$west) tse_filter(., EAST == 1) else . } %>% 
  tse_select(names(vars))

```
# Characteristic
```{r}
tse0 %>%
  tse_meta(rownames = FALSE) %>%
  mytableone(vars)
```
Significant taxa from discovery cohort
```{r}
#df_lm comes from the discovery cohort
sig.list <- df_lm_bmi_results %>%
  dplyr::mutate(across(everything(),~ gsub("GUT_","",.))) %>%
  as.data.frame()
list <- sig.list$taxa
list
#subset tse by sig list
selected_rows <- rowData(tse0)$Species %in% list & 
  !is.na(rowData(tse0)$Species)
tse_val <- tse0[selected_rows, ]
```
#Common taxa
Agglomerate prevalent features
```{r}
#reads were filtered to detection threshold of 0.1% relative abundance, in 5% of samples
#This is to filter out rare taxa, to ensure we are anlaysing microbes that are likely to be real and not sequencing artefacts
subset_features_partial <-  purrr::partial(mia::agglomerateByPrevalence,
                                           x = tse_val,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
Subset to only Species
```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
Melt tse to data frame
```{r}
melt_tse <- function(x, method = "clr") {
    mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltSE(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
## BMI
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_bmi_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_bmi_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_bmi_results %>%
  dplyr::arrange(qval_fdr) %>% 
  DT::datatable(caption = "Linear model for BMI - validation west")
```
Significant taxa from discovery cohort
```{r}
#df_lm comes from the discovery cohort
sig.list <- df_lm_waist_results %>%
  dplyr::mutate(across(everything(),~ gsub("GUT_","",.))) %>%
  as.data.frame()
list <- sig.list$taxa
list
#subset tse by sig list
selected_rows <- rowData(tse1)$Species %in% list &
  !is.na(rowData(tse1)$Species)
tse_val <- tse0[selected_rows, ]
```
#Common taxa
Agglomerate prevalent features
```{r}
#reads were filtered to detection threshold of 0.1% relative abundance, in 5% of samples
#This is to filter out rare taxa, to ensure we are anlaysing microbes that are likely to be real and not sequencing artefacts
subset_features_partial <-  purrr::partial(mia::agglomerateByPrevalence,
                                           x = tse_val,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
Subset to only Species
```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
Melt tse to data frame
```{r}
melt_tse <- function(x, method = "clr") {
    mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltSE(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    #dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
## WaistCirc
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_waist_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_waist_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_waist_results %>%
  dplyr::arrange(qval_fdr) %>%
  DT::datatable(caption = "Linear model for Waist Circ")
```
Significant taxa from discovery cohort
```{r}
#df_lm comes from the discovery cohort
sig.list <- df_lm_whr_results %>%
  dplyr::mutate(across(everything(),~ gsub("GUT_","",.))) %>%
  as.data.frame()
list <- sig.list$taxa
list
#subset tse by sig list
selected_rows <- rowData(tse0)$Species %in% list &
  !is.na(rowData(tse0)$Species)
tse_val <- tse0[selected_rows, ]
```
#Common taxa
Agglomerate prevalent features
```{r}
#reads were filtered to detection threshold of 0.1% relative abundance, in 5% of samples
#This is to filter out rare taxa, to ensure we are anlaysing microbes that are likely to be real and not sequencing artefacts
subset_features_partial <-  purrr::partial(mia::agglomerateByPrevalence,
                                           x = tse_val,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
Subset to only Species
```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
Melt tse to data frame
```{r}
melt_tse <- function(x, method = "clr") {
    mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltSE(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    #dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
## Waist to hip ratio
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using BH
```{r, message=FALSE}
df_lm_whr_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_whr_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::filter(qval_fdr < 0.05)
```
Results
```{r}
df_lm_whr_results %>%
  dplyr::arrange(qval_fdr) %>% 
  #dplyr::filter(qval_fdr < 0.05) %>%
  DT::datatable(caption = "Linear model for Waist Hip ratio")
```